---
title: "Getting started with MORIE in R"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with MORIE in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("morie", quietly = TRUE)
)
```

# Overview

MORIE is a multi-domain scientific-computing toolkit with parallel
Python and R packages. The R package mirrors a substantial subset of
the Python package, focused on the surfaces that are most useful from
within an R workflow: dataset loading, causal estimators, survey
sampling and weighting, basic spectral analysis, and helpers for the
MRM (McNamara--Ruhela--Medina) framework that is MORIE's primary
sociolegal-data application.

This vignette walks through a minimal end-to-end session: load the
package, look at the bundled dataset catalogue, load one dataset, and
run an average-treatment-effect estimator on a small synthetic example.
A second vignette (`mrm-otis-walkthrough`) covers the MRM ten-estimator
ensemble on OTIS provincial data.

# Loading the package

```{r setup}
library(morie)
```

# The dataset catalogue

`morie_dataset_catalog()` returns a data frame summarising every
dataset bundled with the package or accessible via the package's
loaders. This is the easiest way to discover what's available without
leaving the R session.

```{r catalog, eval = FALSE}
catalog <- morie_dataset_catalog()
head(catalog)
```

For details on a single dataset (variables, source, citation), use
`morie_dataset_info()`:

```{r dataset-info, eval = FALSE}
morie_dataset_info("cpads-2122")
```

# Loading a dataset

`morie_load_dataset()` returns a tibble (or data frame) for any
dataset in the catalogue. Public-use datasets that ship inside the
package require no further configuration; for datasets backed by
remote SQLite mirrors, configure `MORIE_LOCAL_DB_DIR` (local
directory of `.sqlite` files) or `MORIE_REMOTE_URL` (HTTP endpoint).

```{r load, eval = FALSE}
df <- morie_load_dataset("cpads-2122")
dim(df)
```

# A simple ATE estimate

For users who already have a treatment / outcome / covariate dataset
in hand, the estimators are designed to work on any tibble or data
frame --- there is no hard-coded column-name convention. The example
below is fully synthetic and runnable without any external data.

```{r ate-synth}
set.seed(2026)
n <- 500
X1 <- rnorm(n)
X2 <- rnorm(n)
# Confounded treatment assignment.
treat <- as.integer(plogis(0.5 * X1 - 0.3 * X2) > runif(n))
# Outcome with a true ATE of +1.0 plus covariate effects.
y <- 1.0 * treat + 0.7 * X1 - 0.2 * X2 + rnorm(n, sd = 0.5)

df_synth <- data.frame(y = y, treat = treat, X1 = X1, X2 = X2)
result <- estimate_ate(
  data       = df_synth,
  outcome    = "y",
  treatment  = "treat",
  covariates = c("X1", "X2")
)
print(result)
```

The returned object is a list with the point estimate, standard error,
confidence interval, and the underlying nuisance fits, in the
`RichResult`-compatible structure described in the Python package
paper.

# Companion estimators

`estimate_att()`, `estimate_atc()`, and `estimate_aipw()` follow the
same calling convention. The augmented IPW estimator
(`estimate_aipw()`) is doubly robust under correct specification of
either the propensity model or the outcome model.

```{r aipw-synth, eval = FALSE}
result_aipw <- estimate_aipw(
  data       = df_synth,
  outcome    = "y",
  treatment  = "treat",
  covariates = c("X1", "X2")
)
print(result_aipw)
```

# Where to go next

- The `mrm-otis-walkthrough` vignette demonstrates the ten-estimator
  MRM ensemble on Ontario OTIS provincial restrictive-confinement
  microdata.
- The MORIE package paper describes the wider scope of the toolkit
  beyond R: signal processing, cryptography, spatial statistics,
  statistical-physics-of-crime models, psychometrics, and the
  full Python interface.
- Citation: see `citation("morie")`.
