---
title: "MRM dataset fetchers and bundled samples"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{MRM dataset fetchers and bundled samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("morie", quietly = TRUE)
)
```

# Overview

The MRM empirical callables (`mrm_otis_*`, `mrm_tps_*`,
`mrm_siu_*`) operate on three external data sources. This vignette
documents how `morie` makes each of them accessible to end users:

* **OTIS** (Offender Tracking Information System; published by the
  Ontario Ministry of the Solicitor General) \u2014 public CKAN
  release on `data.ontario.ca`, reached via the existing
  `morie_load_dataset()` infrastructure.
* **TPS** (Toronto Police Service) \u2014 public ArcGIS Open Data, reached
  via `morie_fetch_tps()`.
* **SIU** (Ontario Special Investigations Unit) \u2014 public Director's
  Reports, scraped on demand by `morie_fetch_siu()` (the corpus is
  not shipped because redistribution licensing is unsettled).

Four small reference samples are bundled with the package
(`inst/extdata/`) so every example runs offline.

```{r intro}
library(morie)
```

# Bundled reference samples

The samples live in `inst/extdata/` and total ~420 KB. Each is a
1000-row random draw (seed 42) except `otis_b09` and `otis_c11`,
which are shipped whole (already small).

```{r samples}
# Bundled samples
b01 <- morie_sample("otis_b01")
b09 <- morie_sample("otis_b09")
c11 <- morie_sample("otis_c11")
tps <- morie_sample("tps_assault")

# Schema sanity
str(b01, max.level = 1)
nrow(b09); nrow(c11)
ncol(tps)
```

# OTIS via CKAN (`data.ontario.ca`)

The OTIS public release is hosted at
`data.ontario.ca/dataset/data-on-inmates-in-ontario` as 28 CSV
resources. The MORIE catalog has four of them registered with their
canonical CKAN resource IDs:

```{r otis-keys}
cat <- morie_dataset_catalog()
cat[cat$source == "otis", c("key", "name", "table_name", "ckan_resource_id")]
```

To pull the full b01 (82,001 rows) from CKAN:

```{r otis-load, eval = FALSE}
b01_full <- morie_load_dataset("otisb01")
nrow(b01_full)
# 82001
```

The loader downloads the CSV on first call, caches it into the
package SQLite database, and returns a data.frame on subsequent
calls.

# TPS via ArcGIS Open Data

Toronto Police Service publishes per-category crime events through
ArcGIS Online. MORIE knows the layer URLs for nine categories.

```{r tps-urls}
morie_tps_layer_urls()
```

To fetch a single category:

```{r tps-fetch, eval = FALSE}
csv_path <- morie_fetch_tps("Assault")
assault <- utils::read.csv(csv_path)
nrow(assault)
# 254378 (as of mid-2026)
```

The fetcher pages through `/query` with a 2000-record-per-page cap
(the ArcGIS-imposed maximum) and writes a tidy CSV to
`~/.cache/morie/tps/tps_Assault.csv`. Subsequent calls return the
cached path unless `overwrite = TRUE`.

Filtering at the server side:

```{r tps-filter, eval = FALSE}
recent <- morie_fetch_tps("Homicides",
                          where = "OCC_YEAR >= 2024",
                          overwrite = TRUE)
```

# SIU via on-demand scraper

The Ontario SIU publishes Director's Reports at
`siu.on.ca/en/case_directors_reports.php`. MORIE includes an
on-demand scraper that:

* Iterates over the index page(s) (optionally filtered by year)
* Extracts case links
* Pulls each case detail page
* Parses incident date, notifying police service, Director's
  decision, and outcome text into a single CSV

```{r siu, eval = FALSE}
csv_path <- morie_fetch_siu()       # full unfiltered index
siu <- utils::read.csv(csv_path)
nrow(siu)

# Or restricted to specific years:
csv_path <- morie_fetch_siu(years = 2020:2025, overwrite = TRUE)
```

## Why a scraper rather than a shipped dataset?

The legal status of redistributing a single tabular copy of public
oversight reports is not clearly established. Running the scraper
per-user is unambiguously fair use of public information; bundling
the scraped corpus might be more questionable. The scraper itself
respects a 2-second rate limit, sets a clear User-Agent, and follows
the SIU site's published structure.

## R-side wrapper, Python-side implementation

`morie_fetch_siu()` in R is a thin `reticulate` wrapper around
`morie.siu_fetch.fetch_siu_cases()` in Python. This keeps the regex
parsing logic in one canonical place. If `reticulate` isn't
installed, fall back to calling the Python directly:

```{bash, eval = FALSE}
python3 -c "from morie.siu_fetch import fetch_siu_cases; print(fetch_siu_cases())"
```

# Workflow: from bundled sample to verified result

Putting it together \u2014 a complete worked example without any network
call:

```{r workflow}
b01 <- morie_sample("otis_b01")

# Mandela classification with default "individual_any" denominator
mrm_classify_mandela(b01, denominator = "row")

# Segregation duration KM
mrm_otis_seg_duration_km(b01,
                         group_cols = "MentalHealth_Alert")

# Mortification co-occurrence: Cramer's V across alert pairs
mrm_otis_mortification_cooccurrence(b01)
```

For a full-data version, swap `morie_sample("otis_b01")` for
`morie_load_dataset("otisb01")` and re-run the same callables.

# Caching layout

| Cache path | Populated by | Size (full) |
|---|---|---|
| `~/.cache/morie/morie.db` (SQLite) | `morie_load_dataset(*)` | a few MB to ~1 GB depending on selection |
| `~/.cache/morie/tps/tps_*.csv` | `morie_fetch_tps()` | ~5--50 MB per category |
| `~/.cache/morie/siu/SIU.csv` | `morie_fetch_siu()` | ~5 MB |

# References

* OTIS data dictionary \u2014
  `data.ontario.ca/dataset/data-on-inmates-in-ontario`
* Toronto Police Open Data \u2014 `data.torontopolice.on.ca/`
* SIU Director's Reports \u2014 `siu.on.ca/en/case_directors_reports.php`
* MRM theoretical paper \u2014 Zenodo DOI
  [10.5281/zenodo.20096075](https://doi.org/10.5281/zenodo.20096075)
