---
title: "MRM empirical callables (OTIS / TPS / SIU)"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{MRM empirical callables (OTIS / TPS / SIU)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("morie", quietly = TRUE)
)
```

# Overview

This vignette documents the `mrm_otis_*()`, `mrm_tps_*()`, and
`mrm_siu_*()` empirical callables. Each function is a one-line entry
point to a verified analysis used in the *MRM empirical paper*
(Ruhela 2026, in preparation). Every example below runs on the small
reference samples bundled with the package, so the vignette is
network-free.

For the full datasets:

* OTIS → `morie_load_dataset("otisb01")` (downloads via CKAN on
  first call; subsequent calls hit the local SQLite cache)
* TPS → `morie_fetch_tps("Assault")` (ArcGIS REST)
* SIU → `morie_fetch_siu()` (on-demand scrape of public reports)

See `vignette("mrm-dataset-fetchers")` for the dataset side.

```{r load}
library(morie)
b01 <- morie_sample("otis_b01")
b09 <- morie_sample("otis_b09")
tps <- morie_sample("tps_assault")
```

# OTIS suite

## Placement-count concentration on `b09`

The b09 long-format file publishes per (fiscal year × placement-count
band × gender) counts of individuals in segregation. The callable
expands the banded counts using midpoints and returns Hill-MLE Pareto
exponent, Gini coefficient, mean placements per individual, and the
top-k% concentration share.

```{r b09}
mrm_otis_placement_concentration(b09)
```

The values are computed *within fiscal year*: OTIS
`UniqueIndividual_ID` has format `YYYY-XXXXX-SG` and is randomly
reassigned every fiscal year, so cross-year tracking is invalid by
design.

## Segregation-duration KM on `b01`

`NumberConsecutiveDays_Segregation` is the duration in days of each
placement (no censoring — all durations are observed). The callable
reports the per-stratum mean, median, q25, and the fraction
above the UN Mandela 15-day cutoff.

```{r b01-duration}
mrm_otis_seg_duration_km(b01)
mrm_otis_seg_duration_km(b01, group_cols = "MentalHealth_Alert")
```

This callable replaces the misreading of `YYYY-XXXXX-SG` as a
persistent person identifier, which produces a spurious cross-year
"time-to-readmission" artifact.

## Mortification co-occurrence (alert columns)

The three b01 alert flags (`MentalHealth_Alert`, `SuicideRisk_Alert`,
`SuicideWatch_Alert`) co-occur to a degree well above independence.
The substantive figure is `MentalHealth × SuicideRisk` Cramer's V.

```{r mortification}
mrm_otis_mortification_cooccurrence(b01)
```

## Region locality

Ontario provincial seg/RC placement is overwhelmingly
locality-preserving — over 95% of placements remain within the same
region in the full b01.

```{r region, eval = FALSE}
# (Region columns are present only in the full b01, not the bundled
# sample; uncomment after morie_load_dataset("otisb01") or
# morie_fetch_tps(...) if needed.)
res <- mrm_otis_region_locality(b01)
print(res$table)
cat("diagonal share:", res$diagonal_share, "  V:", res$cramers_v, "\n")
```

## Mandela classification

`mrm_classify_mandela()` shipped in v0.1.14 and remains the canonical
Mandela classifier in v0.1.15. It supports three operationalisations:

```{r mandela}
mrm_classify_mandela(b01, denominator = "row")           # per-placement
mrm_classify_mandela(b01, denominator = "individual_any") # per-person
mrm_classify_mandela(b01, denominator = "individual_cumulative")
```

The provincial-canonical 12.5/16.5/20.6 % torture rates from c11
require the `c11` aggregate (loaded via `morie_sample("otis_c11")`);
see the MRM empirical paper §6.

# TPS suite

## Levy-flight Hill exponent on inter-event step lengths

Treats consecutive events in chronological order as a single stream
and computes the haversine inter-event step length (km). Returns the
Hill-MLE exponent restricted to steps above `min_step_km`.

```{r tps-levy}
mrm_tps_levy_scaling(tps)
```

## Moran's I + DBSCAN clustering

Grids the WGS84 extent into a coarse raster, counts events per cell,
and computes the global Moran's I via a rook-contiguity matrix. Also
runs DBSCAN on the raw lat/long points (rescaled to km) for cluster
counts.

```{r tps-moran}
mrm_tps_moran_clustering(tps, grid_resolution = 20L)
```

For the high-precision computation on the full 254,378-event Assault
file, use the `morie` Python `tps_spatial_advanced` pipeline; the
R version is for quick interactive auditing.

## Neighbourhood inter-event recurrence

For each `HOOD_158` neighbourhood, sorts events chronologically and
computes the gap (in days) between consecutive events.

```{r tps-recur}
head(mrm_tps_neighbourhood_recurrence_km(tps))
```

## Hawkes manifest loader

`mrm_tps_load_hawkes_refit(path)` reads
`paper_hawkes_refit.json` (the per-category Hawkes refit table from
the MRM empirical paper §7.1-7.2) and returns it as a tidy
data.frame. Skipped here because the JSON is in the MOIRAIS data
manifest, not bundled with the package.

# SIU suite

The SIU callables operate on the SIU.csv file produced by
`morie_fetch_siu()` (an on-demand scraper of the public Director's
Reports). The scraped corpus is not shipped, but the callables
themselves do not depend on shipped data.

```{r siu, eval = FALSE}
siu_path <- morie_fetch_siu()
siu <- read.csv(siu_path)
res <- mrm_siu_case_to_decision_km(siu)
print(res$pooled)
head(res$by_service[order(-res$by_service$n),])
mrm_siu_per_service_rate(siu)
mrm_siu_outcome_classifier(siu)
```

The verified pooled median in our test snapshot is **120 days from
incident to Director's decision** (n = 1,711 cases). Per-service
medians cluster tightly around 120, indicating a system-wide
processing cadence rather than a per-jurisdiction effect.

# References

* MRM theoretical paper — Ruhela (2026), *The McNamara-Ruhela-Medina
  framework*, Zenodo
  [10.5281/zenodo.20096075](https://doi.org/10.5281/zenodo.20096075).
* MRM empirical paper — Ruhela (2026, in preparation),
  `papers/morie-empirical-paper/`.
* OTIS data dictionary — `data.ontario.ca/dataset/data-on-inmates-in-ontario`.
* Toronto Police Open Data — `data.torontopolice.on.ca/`.
* SIU public Director's Reports — `siu.on.ca/en/case_directors_reports.php`.
