---
title: "Signal processing primitives"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Signal processing primitives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("morie", quietly = TRUE)
)
```

# Overview

Beyond its causal-inference and survey-statistics surfaces, MORIE
exposes a small collection of signal-processing primitives. These
are useful for forensic-audio work, biomedical signals, and any
analysis that needs spectral or time-frequency methods adjacent to
the causal pipeline (e.g.\ inter-incident time series in
criminological data).

The R surface is intentionally thin --- the heavy lifting lives in
the Python `morie.signal_processing` and
`morie.homomorphic_deconvolution` modules. The R wrappers cover
the most commonly needed primitives.

# A synthetic two-tone signal

```{r setup}
library(morie)

set.seed(42)
fs <- 1000                           # sampling rate (Hz)
t  <- seq(0, 1, by = 1 / fs)
sig <- sin(2 * pi * 50 * t) +
       0.5 * sin(2 * pi * 120 * t) +
       0.3 * rnorm(length(t))
```

# Basic FFT

```{r fft}
spec <- stats::fft(sig)
n    <- length(sig)
freq <- (0:(n / 2 - 1)) * fs / n
mag  <- Mod(spec)[1:(n / 2)]
peak_freqs <- freq[order(mag, decreasing = TRUE)[1:5]]
peak_freqs
```

The two largest peaks should sit near 50 Hz and 120 Hz, recovering
the synthetic signal's components.

# Where to go next

- Spectral analysis with windowing, multitaper estimation, and
  short-time Fourier transforms is implemented in the Python
  `morie.signal_processing` module --- with the same RichResult
  return convention --- and is documented in the package paper.
- Homomorphic deconvolution and cepstral methods (used in forensic
  audio and biomedical-signal applications) live in
  `morie.homomorphic_deconvolution`.
