---
title: "The SIU pipeline: fetch, parse, audit, translate"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{The SIU pipeline: fetch, parse, audit, translate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("morie", quietly = TRUE)
)
```

# Overview

The Ontario Special Investigations Unit (SIU) publishes director's
reports on every police-civilian interaction that meets the SIU's
mandate (serious injury, death, sexual assault, or discharge of a
firearm at a person). MORIE ships a full pipeline for that corpus:

- a polite-by-default HTTP fetcher (`morie_fetch_siu()`),
- a hand-rolled C++ parser that handles both English and French
  templates from 2005 to the present,
- a language-aware DRID manifest (4,743 drids, English + French) that
  ships inside the package,
- a canonical override system that lets the parser *learn* from
  hand-verified corrections,
- audit + AI tooling (`morie_siu_audit_case()`,
  `morie_siu_anomaly_check()`, `morie_siu_llm_extract()`,
  `morie_siu_compare()`),
- and FR -> EN translation via local Ollama (`morie_siu_translate()`).

Most code chunks below are flagged `eval = FALSE` because they make
network calls or talk to a local LLM. They are tested in `tests/`
under a guarded skipif.

# Inspecting the shipped manifest

The DRID manifest ships inside the package, so reading it requires no
network access.

```{r manifest, eval = FALSE}
library(morie)

manifest <- morie_siu_index()
str(manifest)

# Language counts: en=2531, fr=2212, unknown=0.
table(manifest$`_language`)

# canonical_drid points each French drid at its English counterpart.
head(manifest[manifest$`_language` == "fr",
              c("drid", "case_number", "canonical_drid")])
```

# Fetching the full corpus

`morie_fetch_siu()` walks the manifest, fetches every page (with a
token-bucket throttle, default 4 req/s), and parses the result into a
2,218 x 64 data frame.

```{r fetch, eval = FALSE}
df <- morie_fetch_siu(
  lang       = "en",     # skip French drids: half the network traffic
  cache_html = TRUE,     # persist pages locally; re-runs are ~1-2 min
  rate_limit = 4         # requests per second
)

dim(df)
# 2218 x 64
```

The fetcher respects `Retry-After` on 429 responses, retries up to 3
times with exponential backoff on 5xx, and interleaves report and
news-release fetches in a single rolling-window pool so a cold run
finishes in ~30 minutes rather than the ~58 it took before the
interleaving change in v0.9.5.

# Auditing a single case

`morie_siu_audit_case()` returns the parser row, the raw report HTML,
the raw news HTML, and the cleaned text side-by-side, so you can spot
exactly what the parser saw.

```{r audit, eval = FALSE}
audit <- morie_siu_audit_case("16-OFI-019")

names(audit)
# "row", "report_html", "news_html", "report_text", "news_text"

# Per-field "does the HTML actually support this value?" verdicts.
morie_siu_anomaly_check("16-OFI-019")

# How accurate is each column across a sample of cases?
morie_siu_audit_columns(case_numbers = sample(df$case_number, 50))
```

# Comparing parser output against an external table

`morie_siu_compare()` produces a side-by-side diff between the parser
row and any external table that names the same case.

```{r compare, eval = FALSE}
my_other_table <- data.frame(
  case_number = "16-OFI-019",
  n_officers  = 2L,
  date        = as.Date("2016-01-15")
)

morie_siu_compare(
  case_number = "16-OFI-019",
  external    = my_other_table,
  field_map   = c(officer_count = "n_officers",
                  incident_date = "date")
)
```

# AI extraction (free local model by default)

`morie_siu_llm_extract()` runs an LLM over the cleaned report text and
returns structured field extractions. The default provider is local
Ollama with `gemma3:4b`, so no API key is required.

```{r llm, eval = FALSE}
# Default: local Ollama with gemma3:4b. No API key needed.
morie_siu_llm_extract("16-OFI-019")

# Failover chain: try local first, fall back to Gemini only on error.
morie_siu_llm_extract("16-OFI-019", model = c("ollama", "gemini"))
```

Supported providers: `ollama` (default, local, free), `gemini`,
`claude`, `vertex`. Environment knobs:

- `OLLAMA_HOST` (default `http://localhost:11434`)
- `OLLAMA_MODEL` (default `gemma3:4b`)
- `OLLAMA_KEEP_ALIVE` (default `30m`)
- `OLLAMA_API_KEY` (optional, for hosted gateways)

# French -> English translation

```{r translate, eval = FALSE}
morie_siu_translate(
  text        = "L'enquete a ete close...",
  target_lang = "en"
)
```

`translategemma:latest` is the recommended Ollama model for SIU
French-language report translation; install with `ollama pull
translategemma:latest`.

# The canonical-override system: the parser learns

Ship-time corrections live in
`inst/extdata/siu_canonical_overrides.csv.gz` (47 hand-verified
corrections covering 10 spot-checked cases). Users can add their
own, which take effect on the next `morie_fetch_siu()` call:

```{r override, eval = FALSE}
morie_siu_record_correction(
  case_number = "20-OFD-082",
  field       = "officer_count",
  value       = 3L
)
```

Overrides are applied per cell, by case number, immediately after the
parser writes its initial output.

# Format validity

```{r sanity, eval = FALSE}
sane <- morie_siu_sanity_check(df)
sum(!sane$ok)
# 0 on the shipped corpus (100.000% format-clean)
```

`morie_siu_sanity_check()` validates per-row format constraints:
case-number regex, ISO dates, integer counts, Yes / No values, and
no-page-chrome-leak in narrative columns.

# Where to go next

- `morie_fetch_siu()` is wired through MORIE's universal multi-format
  fetcher; see the `mrm-dataset-fetchers` vignette for the full set
  of supported open-data sources.
- The OTIS provincial corpus is covered in the
  `mrm-otis-walkthrough` vignette; OTIS + SIU together form the
  federal + provincial side of the MRM framework.
- Citation: see `citation("morie")`.
