---
title: "Causal inference with MOIRAIS"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Causal inference with MOIRAIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("moirais", quietly = TRUE)
)
```

# Overview

This vignette walks through the causal-inference surface of MOIRAIS:
ATE, ATT, ATC, augmented IPW (doubly robust), per-unit CATE, group
GATE, and sensitivity analysis (the E-value). Each estimator follows
the same calling convention --- `data`, `treatment`, `outcome`,
`covariates` --- so swapping among them is a matter of changing one
function name.

# A reproducible synthetic example

```{r setup}
library(moirais)
set.seed(2026)

n <- 800
X1 <- rnorm(n)
X2 <- rnorm(n)
# Confounded treatment assignment
ps    <- plogis(0.5 * X1 - 0.3 * X2)
treat <- as.integer(ps > runif(n))
# True ATE = +1.0
y <- 1.0 * treat + 0.7 * X1 - 0.2 * X2 + rnorm(n, sd = 0.5)

df <- data.frame(y = y, treat = treat, X1 = X1, X2 = X2)
```

# Single-robust ATE / ATT / ATC

```{r ate-att-atc}
ate <- estimate_ate(df, treatment = "treat", outcome = "y", covariates = c("X1", "X2"))
att <- estimate_att(df, treatment = "treat", outcome = "y", covariates = c("X1", "X2"))
atc <- estimate_atc(df, treatment = "treat", outcome = "y", covariates = c("X1", "X2"))

ate$estimate
att$estimate
atc$estimate
```

The three quantities estimate slightly different things:

- **ATE** (average treatment effect): averaged over the full
  population.
- **ATT** (treated): averaged over the treated subgroup.
- **ATC** (controls): averaged over the untreated subgroup.

Under no effect heterogeneity in the covariates, all three converge.
Under heterogeneity, they diverge in informative ways.

# Doubly robust: augmented IPW

`estimate_aipw()` is the augmented inverse-probability-weighting
estimator. It is consistent if either the propensity model or the
outcome model is correctly specified --- the doubly-robust
guarantee.

```{r aipw}
aipw <- estimate_aipw(df, treatment = "treat", outcome = "y", covariates = c("X1", "X2"))
aipw$estimate
aipw$se
```

# Per-unit CATE (T-learner / S-learner)

```{r cate}
cate <- estimate_cate(df, treatment = "treat", outcome = "y", covariates = c("X1", "X2"),
                      meta_learner = "t_learner")
head(cate)
```

Each row of the returned data frame contains a per-unit conditional
average treatment effect.

# Group-level GATE

```{r gate}
df$g <- sample(c("A", "B", "C"), nrow(df), replace = TRUE)
gate <- estimate_gate(df, treatment = "treat", outcome = "y",
                      covariates = c("X1", "X2"),
                      group_col = "g")
gate
```

The result is one row per group level with `ate`, `se`, and CI.

# Sensitivity analysis: the E-value

```{r evalue, eval = exists("e_value")}
# Suppose the observed risk ratio is 2.0; how large would an
# unmeasured confounder need to be to explain it away?
evalue <- e_value(rr = 2.0)
evalue
```

The E-value is the minimum strength (on the risk-ratio scale) that
an unmeasured confounder would need on both treatment and outcome
to fully explain the observed effect.

# Where to go next

- The companion `mrm-otis-walkthrough` vignette applies the full
  ten-estimator MRM ensemble to OTIS provincial data.
- For survey-weighted versions of these estimators, see the
  `survey-weighted` vignette.
