---
title: "Survey-weighted estimation with MOIRAIS"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Survey-weighted estimation with MOIRAIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = requireNamespace("moirais", quietly = TRUE)
)
```

# Overview

Many MOIRAIS analyses use Statistics Canada Public Use Microdata
Files (CCS, CSADS, CSUS, CADS, CPADS), which require complex-sample
design weights for valid inference. This vignette covers the
sampling and weighting helpers that wrap the same machinery.

# Stratified, cluster, and PPS sampling

```{r sampling}
library(moirais)
set.seed(11)

n <- 1000
df <- data.frame(
  id     = seq_len(n),
  region = sample(c("ON", "QC", "BC", "AB"), n, replace = TRUE),
  age    = round(rnorm(n, 40, 12)),
  weight = runif(n, 0.5, 2.0)
)

# Stratified random sample within region.
strat <- stratified_sample(df, strata = "region", n_per_stratum = 50)

# Single-stage cluster sample on region (all units in sampled
# clusters are kept).
clust <- cluster_sample(df, cluster = "region", n_clusters = 2)

dim(strat)
dim(clust)
```

# Calibration weights and design weights

`compute_design_weights()` and `calibration_weights()` produce
weights consistent with a known sampling scheme and known
population marginals.

```{r weights, eval = exists("compute_design_weights")}
# Design weights from inclusion probabilities.
design <- compute_design_weights(df, prob = pmin(df$weight / 2, 1))
head(design)
```

# Effective sample size and design effect

```{r ess-deff}
# Effective sample size given a weight vector.
ess <- effective_sample_size(df$weight)
ess

# Design effect: how much variance inflates from the weighting.
deff <- design_effect(df$weight)
deff
```

The effective sample size answers the practical question
"how many independent observations does this weighted sample
contain?" The design effect translates between unweighted-equivalent
and complex-sample variances.

# Bootstrap variance with weights

```{r bootstrap}
# Resample with weights to compute a bootstrap CI.
boot <- bootstrap_sample(df, n_boot = 100, weights = df$weight)
str(boot, max.level = 1)
```

# Where to go next

- For complex-survey causal inference (survey-weighted ATE/AIPW),
  see the `causal-inference` vignette --- the same `estimate_*`
  functions accept a `weights` argument.
- For Statistics Canada PUMF acknowledgments and citation
  requirements, see the package README.
