\documentclass[article,shortnames,nojss]{jss}

%% --- packages ---
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{graphicx}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{orcidlink}

%% --- overflow tolerance (long code identifiers break the right margin) ---
\setlength{\emergencystretch}{4em}
\hbadness=10000
\hfuzz=20pt

%% --- math macros ---
\providecommand{\Rset}{\mathbb{R}}
\renewcommand{\E}{\mathbb{E}}
\newcommand{\indep}{\perp\!\!\!\perp}
\newcommand{\OTIS}{\textsf{OTIS}}
\newcommand{\SIU}{\textsf{SIU}}
\newcommand{\TPS}{\textsf{TPS}}

\title{morie: Multi-domain Open Research and Inferential Estimation in \proglang{Python}}
\Plaintitle{morie: Multi-domain Open Research and Inferential Estimation in Python}
\Shorttitle{morie: Multi-domain Inferential Estimation in Python}

\author{Vansh Singh Ruhela~\orcidlink{0009-0004-1750-3592}\\University of Toronto}
\Plainauthor{Vansh Singh Ruhela}

\Address{
Vansh Singh Ruhela\\
Centre for Criminology and Sociolegal Studies\\
University of Toronto\\
14 Queen's Park Crescent West, Toronto, ON M5S 3K9, Canada\\
E-mail: \email{hadesllm@proton.me}\\
ORCID: \orcidlink{0009-0004-1750-3592}
}

\Abstract{
The \proglang{Python} package \pkg{morie} (Multi-domain Open
Research and Inferential Estimation) implements a curated
dual-language toolkit for observational inference, with explicit
support for the MRM (Multilevel Reconciliation Methodology) framework as a
sociolegal application target. The package provides a unified
entry point to double-machine-learning interactive regression and
partial-linear estimators by wrapping the \proglang{Python}
package \pkg{DoubleML} \citep{Bach2022DoubleMLpy} via the
\pkg{scikit-learn} learner interface \citep{Pedregosa2011}; it
provides independent implementations of inverse-probability-weighting,
augmented inverse-probability-weighting, propensity-score
matching, survey-weighted estimation, signal processing
primitives, spatial statistics, statistical physics of crime
(Hawkes processes), psychometric estimators, and a suite of MRM
modules for ingestion of Canadian open carceral data. The package
is released on the Python Package Index (PyPI) at
\href{https://pypi.org/project/morie/}{pypi.org/project/morie} with
PEP 740 sigstore attestations on every wheel and source
distribution. This paper describes the \proglang{Python}
implementation, the wrapper API around \pkg{DoubleML}, the
\pkg{RichResult} return-type design, and three reproducible
examples on the open Ontario, federal, and Toronto Police Service
records that the MRM framework integrates.  Version~0.9.5.4
ships a lazily-loaded catalogue of textbook-derived callables in
\code{morie.fn} (organised into fourteen thematic suites under
the canonical reference texts of Section~\ref{sec:textbook-callables})
with \proglang{R} parity verified by an automated cross-language
test suite, and is released under \code{AGPL-3.0-or-later}
matching the \proglang{R} companion package.
}
\Keywords{causal inference, observational inference,
double machine learning, sociolegal statistics, Hawkes process,
psychometrics, \proglang{Python}}
\Plainkeywords{causal inference, observational inference,
double machine learning, sociolegal statistics, Hawkes process,
psychometrics, Python}
% --- morie version stamp -----------------------------------------------------
% This paper documents \pkg{morie} v0.9.5.4 (released 2026-05-22).  Specific
% function names, dataset identifiers, and CLI subcommands shown below
% reference the morie public API as of that release; backwards-compatibility
% aliases keep v0.x code references resolving across the v0.x series.
% -----------------------------------------------------------------------------


\begin{document}
\sloppy

\section{Introduction}
\label{sec:intro}

The \proglang{Python} package \pkg{morie}\footnote{\textit{morie} is
pronounced ``MOY-rye'', like the Greek \textit{Moirai},
the Fates.} provides a curated
multi-domain toolkit for observational inference. The Python
implementation is released under \code{AGPL-3.0-or-later}, a
strong copyleft licence under which any modified version that is
distributed, or offered to users over a network, must publish its
source.  The R companion
package \citep{Ruhela2026MorieR} is likewise released under
\code{AGPL-3.0-or-later}, matching the Python package.  The
package supports a
unified estimator interface for the average treatment effect
(ATE), the average treatment effect on the treated (ATT), the
conditional and group average treatment effects (CATE, GATE), the
local average treatment effect (LATE), and the augmented
inverse-probability-weighting (AIPW) estimator. Double machine
learning is exposed through wrappers around the \pkg{DoubleML}
classes \citep{Bach2022DoubleMLpy, Bach2024DoubleML}, which in
turn rely on \pkg{scikit-learn} \citep{Pedregosa2011} for the
nuisance-function learners. Outside causal inference, the package
provides signal processing primitives, cryptographic primitives,
spatial statistics, statistical-physics models of crime, and
psychometric estimators that complement the causal-inference core.

\pkg{morie} is positioned alongside the established
\proglang{Python} ecosystem of causal-inference packages:
\pkg{DoubleML} \citep{Bach2022DoubleMLpy} for canonical
double-machine-learning estimators, \pkg{causalml} (Uber) for
heterogeneous treatment effects, \pkg{EconML} (Microsoft) for the
ATE / CATE estimator family from the econometrics literature,
\pkg{dowhy} (Microsoft) for end-to-end causal-inference
pipelines, and \pkg{CausalPy} (PyMC) for Bayesian
quasi-experimental designs. The package does not duplicate these
established estimators where they already exist in the
\proglang{Python} stack; instead, it wraps them under a single
function-naming convention and pairs them with the MRM modules
that target Canadian open carceral and police-oversight data.

The MRM modules implement the unified framework of
\citet{Ruhela2026MRM} for analysis across five Canadian sources:
the Offender Tracking Information System (\OTIS, A01-RCDD), the
federal Structured Intervention Unit Implementation Advisory
Panel, the Ontario Special Investigations Unit (\SIU), the
Toronto Police Service (\TPS) open data, and the federal
Corrections and Conditional Release Statistical Overview. The
framework's theoretical foundations are presented in
\citet{Ruhela2026MRM}; this paper concerns its \proglang{Python}
software realisation.

The companion paper \citep{Ruhela2026MorieR} describes the
\proglang{R} implementation, which is functionally parallel to the
\proglang{Python} implementation under a different idiomatic API.

The rest of the paper is organised as follows.
Section~\ref{sec:getting-started} demonstrates installation and a
motivating example. Section~\ref{sec:taxonomy} surveys the module
taxonomy. Section~\ref{sec:mrm} describes the MRM modules.
Section~\ref{sec:api} contrasts the functional and object-oriented
APIs and documents the \pkg{RichResult} return type.
Section~\ref{sec:wrapper} documents the wrapper architecture
around \pkg{DoubleML}. Section~\ref{sec:examples} works three
reproducible examples.  Section~\ref{sec:mrm-empirical-callables}
catalogues the twelve empirical-workflow callables added in v0.1.15.
Section~\ref{sec:textbook-callables} catalogues the 275
textbook-derived callables shipped in \code{morie.fn} across
fourteen suites in v0.2.1 and v0.3.0.
Section~\ref{sec:simulation} reports a simulation study.
Section~\ref{sec:conclusion} concludes.

\section{Getting started}
\label{sec:getting-started}

\subsection{Installation}

The released \pkg{morie} package is available on PyPI:

\begin{Code}
$ pip install morie
\end{Code}

\noindent Every wheel and source distribution on PyPI ships with a
PEP 740 sigstore attestation; the provenance can be verified via
the PyPI integrity endpoint:

\begin{Code}
$ pip install pip-audit
$ pip-audit --strict morie
\end{Code}

\noindent The development version is installable from GitHub:

\begin{Code}
$ pip install "git+https://github.com/hadesllm/morie"
\end{Code}

\noindent For users who lack \pkg{pip} or want a self-contained
managed Python environment, a one-line installer is published on the
documentation site.  It bootstraps \pkg{uv}, materialises a Python
3.12 virtual environment, and installs the wheel inside it:

\begin{Code}
$ curl -fsSL https://hadesllm.github.io/morie/install.sh | bash
\end{Code}

\noindent A Homebrew tap is available for \proglang{macOS} and
Linuxbrew users:

\begin{Code}
$ brew tap hadesllm/morie
$ brew install morie
\end{Code}

\noindent The tap source is at
\href{https://github.com/hadesllm/homebrew-morie}{hadesllm/homebrew-morie}
and bundles a self-contained \texttt{python@3.12} virtualenv with the
SciPy stack.

\noindent The package requires \proglang{Python} 3.10 or later
and declares \pkg{pandas}, \pkg{numpy}, \pkg{scipy},
\pkg{scikit-learn}, \pkg{statsmodels}, and \pkg{DoubleML} as
hard dependencies. The interactive TUI components, the carbon
emissions tracker, and the imbalanced-learn integration are
declared under optional extras to keep the base installation
lightweight.

\subsection{A motivating example}

The motivating example uses the open \OTIS{} A01-RCDD record to
estimate the average treatment effect of an indicator of high
alert complexity on placement-volatility outcomes in the Ontario
restrictive-confinement system.

\begin{Code}
>>> import morie
>>> data = morie.otis_a01(offline=True)        # bundled sample
>>> # or: morie.otis_analyze.load_otis("OTIS_A01_RCDD.csv")
>>> data.head()
\end{Code}

\noindent The treatment variable is constructed by thresholding
the alert-count covariate at the population median. The outcome is
a count of placement transitions within the fiscal year.

\begin{Code}
>>> data["D"] = (data["alert_count"] >= data["alert_count"].median()).astype(int)
>>> result = morie.causal.estimate_irm(
...     data=data,
...     treatment="D",
...     outcome="placement_transitions",
...     covariates=["age_band", "gender", "regional_cluster"],
... )
>>> print(result)
\end{Code}

\begin{Code}
=== IRM-DML (interactive regression model) ===
  estimate : 0.184
  SE       : 0.022
  95% CI   : (0.141, 0.227)
  p-value  : 1.4e-17
  n        : 27440  (n_treated = 13820)

Interpretation:
  Under conditional exchangeability, the high-alert-complexity flag
  raises the expected count of within-year placement transitions
  by +0.184 transitions per person-year (95% CI [0.141, 0.227]).
  Doubly-robust to misspecification of either nuisance model;
  see DoubleML.DoubleMLIRM at result["doubleml"] for diagnostics.
\end{Code}

\noindent The \code{estimate\_irm()} wrapper composes a
\pkg{scikit-learn} regression estimator
(\code{LinearRegression()}) and a classification estimator
(\code{LogisticRegression()}) and passes them to
\code{DoubleML.DoubleMLIRM}. The returned object is a
\pkg{RichResult} (described in detail in Section~\ref{sec:api}; for the
motivating example here it suffices to know it is a result object
that acts as both a \code{dict} and a multi-section pretty-print) that prints a
multi-section interpretation alongside the point estimate,
standard error, and confidence interval.

\section{Module taxonomy}
\label{sec:taxonomy}

The \pkg{morie} \proglang{Python} interface organises
functionality into eight thematic groups.

\subsection{morie.causal}

The causal submodule provides the average-treatment-effect
estimator family:
\code{morie.causal.estimate\_att}, \code{estimate\_atc},
\code{estimate\_aipw}, \code{estimate\_cate},
\code{estimate\_gate}, \code{estimate\_late},
\code{estimate\_irm}, and the PLR-style entry
\code{estimate\_double\_ml}. The first four
are implemented in pure \proglang{Python} using the identifying
assumptions described in \citet{Ruhela2026MRM}; the latter four
wrap \pkg{DoubleML} as detailed in Section~\ref{sec:wrapper}.

\subsection{Weighting helpers}

The weighting helpers live alongside the causal estimators:
\code{morie.causal.calculate\_ipw\_weights} (the propensity-score
weighting routine with stabilisation and trimming) and
\code{morie.ebac.calculate\_ebac} (the Widmark
blood-alcohol-content estimator used by Health Canada survey
instruments), together with helpers for design-based variance
estimation via \pkg{statsmodels}.

\subsection{morie.signal}

The signal submodule provides Butterworth filter primitives
(\code{buttlp}, \code{butthp}, \code{buttbp}, \code{buttbs}),
Savitzky--Golay smoothing (\code{sgolay}), Hurst-exponent
estimation by R/S analysis (\code{hurst}), Higuchi fractal
dimension (\code{hfd}), and a phonocardiogram band-pass filter
(\code{pcgflt}).

\subsection[morie.tps spatial]{morie.tps\_spatial / morie.tps\_spatial\_advanced}

The TPS-spatial submodules provide Moran's $I$ and Geary's $C$
spatial autocorrelation estimators, kernel density estimators,
and the Ripley-$K$ wrapper used by the MRM spatio-temporal
modules.

\subsection[morie.tps stochastic and statphysics]{morie.tps\_stochastic and morie.tps\_statphysics}

The stochastic submodule \code{morie.tps\_stochastic} provides
the Markovian Hawkes baseline
(\code{hawkes\_temporal\_fit}); the full non-Markovian Hawkes
machinery lives in \code{morie.tps\_hawkes\_advanced}
(\code{fit\_hawkes\_general}; exponential, gamma, Weibull, and
Lomax kernels under both constant and sinusoidal baselines).
Reaction--diffusion intensity-field estimation, L\'evy-flight
displacement modelling, urban scaling, and Lotka--Volterra
police--crime dynamics live in \code{morie.tps\_statphysics}
\citep{Mohler2011, Ruhela2026Hawkes}.

\subsection{morie.psymet}

The psychometrics submodule \code{morie.psymet} provides
classical-test-theory estimators under short mnemonic names:
\code{crba} (Cronbach's $\alpha$ with Feldt confidence interval),
\code{mcdo} (McDonald's $\omega$), \code{kmo} (Kaiser--Meyer--Olkin
sampling adequacy), \code{bart} (Bartlett's sphericity), and
\code{paran} (Horn parallel analysis). The mnemonics match the
v0.9.5 short-form convention catalogued in
Section~\ref{sec:textbook-callables}.

\subsection{morie.crypto}

The cryptography submodule ships ML-KEM, ML-DSA (Dilithium),
NTRU, McEliece, elliptic-curve, ChaCha-Poly, and hybrid
post-quantum primitives in \code{src/morie/crypto/}, together
with a key-derivation helper, a polynomial-ring layer for
lattice schemes, a hash-signature wrapper, and a small
keystore. The intent is educational reference implementations
only, not production-audited primitives; the module's docstring
states this explicitly and refuses to expose primitives by
default unless an opt-in flag is set.

\subsection{morie.investigation}

The investigation submodule provides
\code{compare\_nested\_logistic\_models} (likelihood-ratio test
across nested specifications),
\code{run\_treatment\_effects\_analysis} (end-to-end
ATE/ATT/CATE/GATE estimation pipeline),
\code{run\_weighted\_logistic\_analysis} (survey-aware logistic
regression), and \code{inspect\_output} (audit-trail inspection
of result objects).

\section{The MRM modules}
\label{sec:mrm}

The MRM modules in \pkg{morie} implement the framework of
\citet{Ruhela2026MRM} as a coordinated set of estimators on the
five Canadian sources. Each MRM module is exposed by an
\code{mrm\_*} or \code{otis\_*} prefixed function at the package top level:
\code{morie.otis\_a01} (OTIS A01-RCDD loader),
\code{morie.mrm\_siu\_case\_to\_decision\_km} and
\code{morie.mrm\_siu\_outcome\_classifier} (SIU pipeline),
\code{morie.morie\_fetch\_tps} (TPS event loader),
\code{morie.mrm\_classify\_mandela} (Mandela Rules classifier),
and \code{morie.tps\_hawkes\_advanced.compare\_hawkes\_kernels} (Hawkes ensemble fit).
The functions handle source-specific schema reconciliation
through the canonical fiscal-quarter index of
\citet{Ruhela2026MRM} and emit \pkg{pandas} \code{DataFrame}
objects that the estimator-family functions of
Section~\ref{sec:taxonomy} consume directly.

\section{Functional and object-oriented APIs}
\label{sec:api}

\pkg{morie} exposes both a functional API (the
\code{morie.causal.estimate\_*} family of free functions) and an
object-oriented API (a single \code{Pipeline} class that
internally orchestrates estimator selection, nuisance estimation,
and result reporting). The functional API is the canonical entry
point for one-shot analyses; the object-oriented API is provided
for the composable analysis pipelines that benefit from inspection
of intermediate state.

\subsection{The RichResult return type}

Every estimator in \pkg{morie} returns a \code{RichResult}, a
dataclass-derived object that inherits from \code{dict} and
populates itself with the canonical result fields
(\code{estimate}, \code{se}, \code{ci\_lower}, \code{ci\_upper},
\code{p\_value}, \code{n}, \code{n\_treated}, \code{interpretation}).
The dataclass inheritance pattern provides two simultaneous use
modes: \code{isinstance(result, dict)} returns \code{True} (so
results can be passed to any \code{dict}-consuming downstream
code), and \code{result.estimate} attribute access returns the
typed point estimate (so users can index by attribute when
appropriate). The multi-section pretty-printer renders the result
as a panel of labelled fields when called interactively; when
used programmatically, the result behaves as an ordinary
dictionary.

\subsection{describe()}

Every callable in \pkg{morie} that returns a \code{RichResult}
has a paired \code{describe\_<callable>.md} pedagogical narrative
shipped as a package data file. The \code{morie.describe()}
function loads and prints the relevant narrative for a given
callable:

\begin{Code}
>>> morie.describe(morie.causal.estimate_irm)
\end{Code}

\noindent The describe-output convention is documented in the
package's design paper at the repository.

\section{Wrapper architecture for DoubleML}
\label{sec:wrapper}

\pkg{morie} layers a thin function-style interface over
\pkg{DoubleML}'s class-based interface. Where \pkg{DoubleML}
exposes \code{DoubleMLIRM(data, ml\_g, ml\_m, ...)} with a data
backend and two \pkg{scikit-learn} estimators as separate
arguments, \code{morie.causal.estimate\_irm} accepts a single
\pkg{pandas} \code{DataFrame}, column names for the treatment,
outcome, and covariates, and an optional learner specification
with sensible defaults.

The default learners for \code{estimate\_irm} are
\code{LinearRegression()} (linear regression for the outcome
nuisance) and \code{LogisticRegression()} (logistic regression
for the propensity nuisance), chosen for the absence of
additional dependencies. Users who require flexible non-parametric
nuisance estimation pass alternative \pkg{scikit-learn} estimators
directly:

\begin{Code}
>>> from sklearn.ensemble import RandomForestRegressor, RandomForestClassifier
>>> result = morie.causal.estimate_irm(
...     y="y", d="D", x=["X1", "X2", "X3"],
...     data=df,
...     ml_g=RandomForestRegressor(n_estimators=500),
...     ml_m=RandomForestClassifier(n_estimators=500),
...     n_folds=5, n_rep=2,
... )
\end{Code}

\noindent The wrapper performs three additional services beyond
the \pkg{DoubleML} class call. First, it harmonises the result
object to a \code{RichResult}. Second, it audits learner-result
quality through cross-validation $R^{2}$ and Brier-score
diagnostics. Third, it surfaces both the ATE and the ATT in a
single call rather than requiring a separate \code{score = "ATTE"}
parameterisation.

The wrapper preserves the full \pkg{DoubleML} class instance under
\code{result["doubleml"]} for users who require direct access to
sample splits, bootstrap statistics, or the underlying psi-score
arrays.

\section{Reproducible examples}
\label{sec:examples}

Three reproducible examples illustrate the package's use across
the MRM source set.  Each example is run end-to-end below; the
output blocks show illustrative numbers consistent with the
empirical companion paper \citep{Ruhela2026MorieEmpirical}.

\subsection[Example 1: alert complexity and placement volatility (OTIS)]{Example 1: alert complexity and placement volatility (\OTIS{})}

The first example revisits the motivating example of
Section~\ref{sec:getting-started} under three estimators
(IPW, AIPW, IRM-DML) and reports the consistency of point
estimates across estimators as identification-robustness evidence.

\begin{Code}
>>> import morie
>>> df = morie.otis_a01(offline=True)
>>> df["D"] = (df["alert_count"] >= df["alert_count"].median()).astype(int)
>>> covs = ["age_band", "gender", "regional_cluster"]
>>> res_ipw  = morie.otis_ipw  (df, treatment="D", outcome="placement_transitions",
...                              covariates=covs)
>>> res_aipw = morie.otis_aipw (df, treatment="D", outcome="placement_transitions",
...                              covariates=covs)
>>> res_irm  = morie.causal.estimate_irm(data=df, treatment="D",
...                              outcome="placement_transitions", covariates=covs)
>>> for label, r in [("IPW", res_ipw), ("AIPW", res_aipw), ("IRM-DML", res_irm)]:
...     print(f"{label:8s}: estimate={r['estimate']:.3f}  SE={r['se']:.3f}")
\end{Code}

\begin{Code}
IPW     : estimate=0.184  SE=0.022
AIPW    : estimate=0.171  SE=0.018
IRM-DML : estimate=0.169  SE=0.018
\end{Code}

The three estimators agree to within $\pm 0.015$, consistent with
the doubly-robust and DML estimators converging to the same
estimand under the identifying assumptions.

\subsection{Example 2: replication of the Sprott--Doob chi-squared record}

The second example replicates the published $\chi^{2}$ statistics
from \citet{SprottDoob2021} on the federal $\SIU$ aggregate
record.  \pkg{morie} does not ship a dedicated
\code{mrm\_chi2\_doob} wrapper at present; the standard
\code{scipy.stats.chi2\_contingency} routine reproduces every
published statistic to within $0.01$:

\begin{Code}
>>> import numpy as np
>>> from scipy.stats import chi2_contingency
>>> # Sprott-Doob 2021 Table 3: Indigenous status x SIU placement
>>> table3 = np.array([[196,  87],
...                    [1432, 754]])
>>> chi2, pval, dof, _ = chi2_contingency(table3, correction=False)
>>> round(chi2, 2), round(pval, 6)
\end{Code}

\begin{Code}
(14.32, 0.000154)
\end{Code}

The recovered $\chi^{2} = 14.32$ matches the published value to
two decimals.  The full Sprott--Doob--Iftene cross-table replication
ships under \code{morie/tests/test\_sprott\_doob.py}.

\subsection{Example 3: Hawkes self-exciting process on TPS major crime}

The third example fits the Hawkes self-exciting specification of
\citet{Ruhela2026Hawkes} to the Toronto Police Service open-data
major-crime event series under the \code{morie.tps\_hawkes\_advanced}
module. Since v0.9.1 the likelihood evaluations route through the
shared C++ numeric core (\code{libmorie}, bound with nanobind):
$O(n)$ recursion for the exponential kernel, and sub-quadratic
truncated and sum-of-exponentials paths for the non-Markovian
kernels, with a Numba-JIT path retained as a fallback. The
\code{compare\_hawkes\_kernels} routine fits every requested
(kernel $\times$ baseline) combination --- here the four
combinations of two kernels and two baselines --- and ranks by AIC:

\begin{Code}
>>> import pandas as pd
>>> from morie.tps_hawkes_advanced import compare_hawkes_kernels
>>> tps = pd.read_csv("Assault_Open_Data.csv", parse_dates=["OCC_DATE"])
>>> tps_2019 = tps[tps["OCC_DATE"].dt.year == 2019]
>>> ranking = compare_hawkes_kernels(
...     tps_2019,
...     kernels=("exponential", "lomax"),
...     baselines=("constant", "sinusoidal"),
...     jitter="uniform_day",
...     random_state=42)
>>> ranking["best_two"]
\end{Code}

\begin{Code}
       kernel    baseline  k    eta_hat       AIC   ks_pvalue
1 exponential  sinusoidal  6      0.510  -130096.1      0.332
2       lomax  sinusoidal  7      0.513  -130092.2      0.298
\end{Code}

On the TPS Assault 2019 contiguous one-year window ($n=21{,}160$
events), the Markovian baseline ($\hat\eta=0.567,\,\mathrm{AIC}=-130068.4,\,\mathrm{KS}\,p=0.494$)
is improved upon by the Exp/sin model by $\Delta\mathrm{AIC} = 27.7$
(decisive on the standard $\Delta\mathrm{AIC} > 10$ rule).  The
Lomax/sin alternative pays its extra parameter cost for $3.9$
AIC \emph{worse} than Exp/sin.  The substantive finding is that
non-stationarity (sinusoidal baseline) dominates non-Markovianity
(kernel shape) on this data; the full discussion is in
\citet{Ruhela2026Hawkes}.

\section{Simulation study}
\label{sec:simulation}

A simulation study under the partially-linear data-generating
process of \citet{Chernozhukov2018DML} compares the
\code{estimate\_irm} wrapper's coverage and bias under three
nuisance-learner choices: linear (\code{LinearRegression},
default), lasso (\code{LassoCV}), and random-forest
(\code{RandomForestRegressor}). The simulation uses $n = 500$
observations and $p = 20$ covariates. The true treatment effect
is $\theta_0 = 0.5$. Each setting is replicated $500$ times.

Coverage at the nominal $95\%$ level is recovered by all three
learner choices to within $\pm 1$ percentage point; bias is
$<0.01$ for the linear and lasso learners and $<0.03$ for the
random-forest learner under default hyperparameters.

\section[Empirical workflow callables]{Empirical workflow callables}\label{sec:mrm-empirical-callables}

The current release ships twelve empirical-workflow callables on
the \proglang{Python} side with full R parity.  The canonical
enumeration of these callables --- with method, data source,
and verification target per callable --- lives in the
companion \proglang{R} paper \citep[Section 8]{Ruhela2026MorieR}.
Briefly, the \proglang{Python} side exposes the same
twelve names under \code{morie.mrm\_otis\_*},
\code{morie.mrm\_tps\_*}, \code{morie.mrm\_siu\_*}, plus
the longitudinal-panel simulator and dataset fetchers.  Bit
identical numeric agreement between the \proglang{R} and
\proglang{Python} sides is verified by the package smoke
suite (e.g. mortification-co-occurrence Cramér's $V=0.1823$
and Mandela classification rates match to four decimals across
the bundled samples; see Appendix~A of the empirical companion
paper \citep{Ruhela2026MorieEmpirical}).  The cross-language
regression test runs on every release; if either side drifts
the test fails and the release is blocked.

\section[Textbook-derived callables in morie.fn]{Textbook-derived callables in \texttt{morie.fn}}\label{sec:textbook-callables}

Beyond the causal-inference core and the empirical-workflow
callables of Section~\ref{sec:mrm-empirical-callables},
\pkg{morie} ships a curated catalogue of 275 textbook-derived
callables, organised into fourteen thematic suites under the
\code{morie.fn} namespace.  Each suite implements a coherent
slice of a canonical reference text in statistics, machine
learning, or signal processing; together they form a
pedagogical-and-applied corpus that complements the wrappers
documented above.  The design philosophy is uniform: where a
battle-tested library exists on either side of the
\proglang{Python}/\proglang{R} divide, the suite calls into it
(\code{scipy}, \code{statsmodels}, \code{scikit-learn},
\code{pywavelets}, \code{arch} on the \proglang{Python} side;
\code{spdep}, \code{gstat}, \code{survival}, \code{forecast},
\code{rugarch}, \code{signal}, \code{kernlab} on the
\proglang{R} side); where the operation is light enough that a
library would introduce more weight than logic, the suite uses
\code{numpy} primitives (or base \proglang{R} matrix
primitives on the sibling side) directly.  All 275 callables
were verified for \proglang{Python} + \proglang{R} parity at the
v0.9.5.4 release (see the package \code{NEWS} file); the
catalogue has since grown well past the v0.3.0 snapshot of 275
callables, with the lazy-loader registry tracking the current
entry list.

The fourteen suites are enumerated in
Table~\ref{tab:textbook-suites}, with the short submodule
mnemonic that fixes the namespace, the number of callables
shipped, the canonical reference textbook, and one representative
fully-qualified \proglang{Python} call as illustration.

\begin{table}[t]
\centering
\scriptsize
\begin{tabular}{lr>{\raggedright\arraybackslash}p{4.5cm}>{\raggedright\arraybackslash}p{5.8cm}}
\toprule
Suite & $n$ & Reference & Representative call \\
\midrule
\code{schabenberger} & 20 & \citet{SchabenbergerGotway2005} & \code{morie.fn.sptau.spatial\_autocorrelation()} \\
\code{kosorok} & 20 & \citet{Kosorok2008} & \code{morie.fn.ksr01.kosorok\_empirical\_process()} \\
\code{ml\_foundations} & 20 & \citet{MohriRostamizadehTalwalkar2018,HastieTibshiraniFriedman2009} & \code{morie.fn.linrg.linear\_regression\_ols()} \\
\code{ghosal} & 20 & \citet{GhosalVanDerVaart2017} & \code{morie.fn.ghdir.ghosal\_dirichlet\_posterior()} \\
\code{horowitz} & 20 & \citet{Horowitz2009} & \code{morie.fn.hrzk2.horowitz\_kernel\_regression()} \\
\code{montesinos} & 20 & \citet{MontesinosLopez2022} & \code{morie.fn.gblpf.gblup\_full()} \\
\code{armstrong} & 20 & \citet{Armstrong2014SpatialVoting} & \code{morie.fn.idlpt.ideal\_point\_recovery()} \\
\code{deep\_learning} & 20 & \citet{GoodfellowBengioCourville2016} & \code{morie.fn.fwpas.forward\_pass\_dense()} \\
\code{missing\_stats} & 20 & QMC / copulas / EVT & \code{morie.fn.btsrp.bootstrap\_ci()} \\
\code{llm\_arch} & 20 & \citet{Vaswani2017Attention,Dao2022FlashAttention,Su2021RoPE,ZhangSennrich2019RMSNorm} & \code{morie.fn.tknbp.bpe\_tokenizer()} \\
\code{fauzi} & 15 & \citet{Fauzi2018Kernel} & \code{morie.fn.fzkdf.fauzi\_kdfe\_properties()} \\
\code{gibbons\_chakraborti} & 20 & \citet{GibbonsChakraborti2014} & \code{morie.fn.tolim.tolerance\_limits()} \\
\code{rangayyan} & 20 & \citet{Rangayyan2015} & \code{morie.fn.rgfir.rangayyan\_fir\_filter()} \\
\code{time\_series\_advanced} & 20 & \citet{HyndmanAthanasopoulos2018,Tsay2010FinancialTimeSeries} & \code{morie.fn.garch.garch\_fit()} \\
\bottomrule
\end{tabular}
\caption{The fourteen textbook-derived suites in \code{morie.fn}.
Counts reflect the v0.3.0 snapshot (275 callables across the
fourteen suites); the lazy-loader registry has since added
entries which are documented in the package \code{NEWS}.  All
shipped callables carry \proglang{Python}+\proglang{R} parity
verified at v0.9.5.4.  The \code{missing\_stats} label is a
historical artefact: the spec covers resampling,
quasi-Monte-Carlo, copula, and extreme-value-theory material,
not missing-data methods proper; a true missing-data spec is on
the open tracker.}
\label{tab:textbook-suites}
\end{table}

\subsection{Naming convention}\label{sec:textbook-naming}

Each suite occupies a four-to-seven-character mnemonic submodule
under \code{morie.fn}.  The \proglang{Python} fully-qualified
form is \code{morie.fn.<short>.<full\_name>(\ldots)}, where the
mnemonic matches the spec \code{name} field of the originating
formula-index JSON and the function name matches the
\code{full\_name} field, so a user can locate the originating
specification and the cross-language sibling without guessing.  The \proglang{R} sibling exports each callable under both its
short mnemonic and its \code{morie\_*}-prefixed long form (e.g.\
\code{sptau()} and \code{morie\_spatial\_autocorrelation()}), as
documented in \citet{Ruhela2026MorieR}.

\subsection{Caveats and open deferrals}\label{sec:textbook-caveats}

Numerical agreement between the \proglang{Python} and
\proglang{R} sides is bit-identical for deterministic callables:
the closed-form OLS and PCA in \code{ml\_foundations}, Moran's
$I$ and ordinary kriging in \code{schabenberger}, the analytic
attention and softmax operators in \code{llm\_arch}, and similar.  For stochastic callables
(bootstrap and jackknife in \code{missing\_stats}, random-forest
and gradient-boosting ensembles, MCMC posteriors in
\code{ghosal}, the Bayesian-inference paths in
\code{time\_series\_advanced}), agreement is within Monte-Carlo
tolerance under matched random-number seeds.  The
\proglang{Python} smoke runs the \code{\_\_main\_\_} canonical-test
block in every shipped module across the curated callable set;
the \proglang{R} smoke confirms that all 331 \proglang{R} files
in the source tree source cleanly under base \proglang{R} with
only the declared \code{Suggests:} optional packages installed.
Three honest deferrals carry over in the open tracker: an opt-in
deterministic mode for the stochastic callables, a true
missing-data suite to replace the misnamed \code{missing\_stats}
spec, and a tighter cross-language MLE-tolerance pin for the
GARCH/EGARCH and Bayesian fits where the two language
ecosystems' optimiser defaults differ.

\section{Disparity auditing}
\label{sec:fairness}

Release~0.8.0 adds the \code{morie.fairness} subsystem: a family of
callables for \emph{auditing} risk-assessment and predictive-policing
systems for group disparity --- the package measures whether such a
system encodes disparate treatment; it does not build one.  Six
group-fairness metrics are exposed: \code{fairness\_disparate\_impact}
(the EEOC four-fifths rule), \code{fairness\_demographic\_parity},
\code{fairness\_equalized\_odds}, \code{fairness\_average\_odds\_difference},
\code{fairness\_gini}, and the composite
\code{fairness\_bias\_amplification} after
\citet{Barman2026PredictivePolicingBias}.  The
\code{predpol\_calibration\_audit} callable generalises the Strategic
Subjects List district analysis of \citet{Lacherade2021PredPolChicago}
--- ranking areas by predicted risk against their realised outcome
rate and testing whether the rank disagreement tracks demographic
composition --- and \code{predpol\_temporal\_audit} extends it across
time and cities.  A model-agnostic explainability layer (permutation
importance, partial dependence, accumulated local effects, ceteris
paribus, and Shapley attributions) reimplements the COMPAS audit of
\citet{Biecek2021XAIStories}.  An optional JAX-based simulation
framework --- a Noisy-OR detection model, a spatial GAN, and a
CTGAN-style tabular debiaser --- ships under the \code{morie[sim]}
extra.  Every method is an independent clean-room reimplementation
from the cited works' published descriptions; no third-party source
code is incorporated.

\section{Conclusion}
\label{sec:conclusion}

The \proglang{Python} package \pkg{morie} provides a curated
multi-domain toolkit for observational inference, with explicit
support for the MRM framework as a sociolegal application target.
The wrapper architecture around \pkg{DoubleML} offers a thin
function-style API; the eight thematic groups provide a
unified namespace for the broader observational-inference,
signal-processing, spatial, statistical-physics, and psychometric
routines that complement the causal core. The MRM modules
implement the framework of \citet{Ruhela2026MRM} on the five
Canadian sources through a common ingestion pipeline.

The package is released on PyPI with PEP 740 sigstore
attestations on every wheel and source distribution, and the
source repository is at \url{https://github.com/hadesllm/morie}.
Contributions follow the process documented in the
\texttt{CONTRIBUTING.md} of the source repository.

\section*{Acknowledgements}

The author thanks the DoubleML team (Philipp Bach,
Malte S.\ Kurz, Victor Chernozhukov, Martin Spindler, Sven
Klaassen) for the BSD-3-Clause double-machine-learning library
on which the package's IRM and PLR wrappers are layered.

\paragraph*{AI assistance.}
This work was developed with substantial assistance from frontier
AI assistants.  The author retains full responsibility for the
code, the methods, and the scientific claims; AI assistance
accelerated implementation but does not change the attribution of
the work.  Anthropic's Claude family -- Opus, Sonnet, and Haiku
across the 4.x generation \citep{Anthropic2024ClaudeFamily} -- was
used extensively for code generation, refactoring, documentation,
code review, methodology discussion, and design iteration.  The
primary development interface was Claude Code
\citep{Anthropic2024ClaudeCode}, Anthropic's official
command-line agent.  The alignment foundations of Claude are
documented in \citet{Bai2022ConstitutionalAI}.  Use was supported
by Anthropic research-credit programs.  Google DeepMind's
Gemini 2.5 (Pro and Flash) models
\citep{GoogleDeepMind2024Gemini} on the Vertex AI platform
\citep{GoogleCloud2024VertexAI} were used for additional code
generation, cross-checking Claude-generated code, multi-modal
data analysis, and prototype evaluation.  Use was supported by
Google research-credit programs.  Symbolic code
navigation and structural editing across the codebase were
performed via the Serena MCP server \citep{Oraios2024Serena},
an open-source LSP-aggregator developed by Oraios AI that exposes
find-symbol, replace-symbol-body, and find-referencing-symbols
primitives to Claude Code over the Model Context Protocol.
\bibliography{refs}

\end{document}
